# GitHub

## GitHub Introduction lecture

Here's a recording of this lecture (32 minutes 8 seconds):

[Recording](https://pitt.hosted.panopto.com/Panopto/Pages/Viewer.aspx?id=e274c96f-1906-4955-9a8d-b06600f51f49){.external target="_blank"}

<iframe credentialless src="https://pitt.hosted.panopto.com/Panopto/Pages/Embed.aspx?id=e274c96f-1906-4955-9a8d-b06600f51f49&amp;autoplay=false&amp;offerviewer=true&amp;showtitle=true&amp;showbrand=true&amp;captions=true&amp;interactivity=all" height="405" width="720" style="border: 1px solid #464646;" allowfullscreen allow="autoplay">

</iframe>

## GitHub Introduction slides

[PDF slide set](pdfs/GitHubIntro.pdf){target="_blank"}

<iframe width="100%" height="800" src="pdfs/GitHubIntro.pdf">
</iframe>

## Questions

### Question 1

Which of the following benefits does version control offer over emailing files back and forth?

Please select the correct options (more than one may be correct). 

```{r, echo=FALSE}
library(webexercises)
choices = c(
answer="It provides a complete history of file changes, including who made them and when",
"It eliminates the need for collaboration on projects",
answer = "It prevents accidental overwrites and lost files",
"It automatically detects and resolves all conflicts between users",
answer = "It makes it easy to revert to previous versions of files"
)
```

`r longmcq(choices)`

### Question 2

Which Git command is used to obtain a complete copy of a remote repository for the first time?

Please select the correct option. 

```{r, echo=FALSE}
library(webexercises)
choices = c(
"git pull",
"git fetch",
answer = "git clone",
"git push"
)
```

`r longmcq(choices)`

 

### Question 3

Which of the following Git workflows is generally considered best practice for efficient and collaborative development?

Please select the correct option.  

```{r, echo=FALSE}
library(webexercises)
choices = c(
"Pull, commit, push",
answer = "Pull, work, commit, pull, push",
"Commit, push, pull",
"Pull, work, commit, push"
)
```

`r longmcq(choices)`


