# Genetic Data in R and GDS

## Genetic Data in R and GDS lecture

Here's a recording of this lecture (50 minutes 56 seconds):


[Recording](https://pitt.hosted.panopto.com/Panopto/Pages/Viewer.aspx?id=bda73d86-08f7-4159-876c-b2310136c45b){.external target="_blank"}

<iframe credentialless src="https://pitt.hosted.panopto.com/Panopto/Pages/Embed.aspx?id=bda73d86-08f7-4159-876c-b2310136c45b&amp;autoplay=false&amp;offerviewer=true&amp;showtitle=true&amp;showbrand=true&amp;captions=true&amp;interactivity=all" height="405" width="720" style="border: 1px solid #464646;" allowfullscreen allow="autoplay">

</iframe>

## Genetic Data in R and GDS slides

[PDF slide set](pdfs/genetic_data_in_r_gds_lecture){target="_blank"}

<iframe width="100%" height="800" src="pdfs/genetic_data_in_r_gds_lecture">

</iframe>

## Questions

### Question 1

In the file tree on slide 7, if your current directory is backup (inside of thing), how could you change your directory to the other folder with the same name?

Please select the correct options (more than one may be correct).

```{r, echo=FALSE}
library(webexercises)
choices = c(
answer="cd /Users/backup",
"cd backup",
"pwd",
"cd",
answer = "cd ../../backup"
)
```

`r longmcq(choices)`

### Question 2

If your command results in an error message about a file not being found, what should you probably do first?

Please select the correct option.

```{r, echo=FALSE}
library(webexercises)
choices = c(
"cd to your home directory",
answer = "pwd to make sure you're in the correct directory",
"cd / to go to the root directory",
"git push"
)
```

`r longmcq(choices)`

### Question 3

What will happen if you enter the command `cp file1 file2` but `file2` is a file that already exists?

Please select the correct option.

```{r, echo=FALSE}
library(webexercises)
choices = c(
"there will be an error message that prevents you from overwriting the existing file2",
"a new copy of file1 named file2_copy(1) will be created",
"file1 will be renamed as file2",
answer="the original file2 will be permanently overwritten with a new copy of file1"
)
```

`r longmcq(choices)`

### Question 4

What command will set the file permissions so that you, the owner/user, have `rwx` permissions without changing other users' permissions?

Please select the correct option.

```{r, echo=FALSE}
library(webexercises)
choices = c(
answer="chmod -v file u=rwx file",
"chmod -v file o=rwx file",
"chmod -v file a=rwx file",
"chmod -v file u+x file"
)
```

`r longmcq(choices)`

### Question 5

What is the difference between `tail -n20 file` and `tail -n+20 file`?

Please select the correct option.

```{r, echo=FALSE}
library(webexercises)
choices = c(
"The first one prints the last 20 lines, but the second one prints the first 20 lines.",
"There is no difference.",
answer="The first one prints the last 20 lines, but the second one prints from line 20 to the end of the file.",
"The first one prints the last 20 lines, but the second one just prints line 20."
)
```

`r longmcq(choices)`

### Question 6

What git commands will you use to submit your homework (assume your working directory is your homework repository)?

Please select the correct option.

```{r, echo=FALSE}
library(webexercises)
choices = c(
"push",
"git add and then git push",
answer="First do git pull to make sure your repo is up to date. Then use git add to add the files you want to upload. Then git commit -m with a message. Then git push.",
"Just git push."
)
```

`r longmcq(choices)`

### Question 7

How do you print out the exit status of the last process?

Please select the correct option.

```{r, echo=FALSE}
library(webexercises)
choices = c(
"echo 0",
"echo exit_status",
"just type a dollar sign and then a question mark",
answer="echo $?"
)
```

`r longmcq(choices)`

### Question 8

What is the first thing you should do after logging on to the computing cluster?

Please select the correct option.

```{r, echo=FALSE}
library(webexercises)
choices = c(
answer="srun -M teach -A hugen2071-2024f --pty bash",
"pwd",
"git pull",
"exit"
)
```

`r longmcq(choices)`