# R Graphics Exercise

```{r,echo=FALSE,message=FALSE,warning=FALSE}
require(knitr)
# Set so that long lines in R will be wrapped:
# opts_chunk$set(tidy.opts=list(width.cutoff=80),tidy=TRUE)
```

## Load Libraries

```{r load_libraries,message=FALSE}
library(tidyverse)
library(ggforce)
# library(tidylog)
# Set the default font to be a bit larger:
theme_set(theme_gray(base_size = 18))
```

## Exercise 1

Read in and set up the data set `b`, a cleaned version of our simulated data set:

```{r}
a <- read.csv("data/study1.csv")
a$ind <- seq_along(a$t)
b <- a[-c(1001:1004),]
b$g.f <- factor(b$g)
b$geno <- paste(b$all1,b$all2,sep="/")
```

Using ggplot and data set `b`, plot `ind` vs. `t`, coloring by case-control status (`trait`). What do you observe about the data?

::: {.callout-tip collapse="true"}
## Expand to see solution

```{r}
ggplot(data=b, aes(x=ind, y=t, color=trait)) + 
   geom_point()
```
:::

## Exercise 2

Using ggplot, plot `ind` vs. `t`, coloring by case-control status (`trait`) and faceting by `geno`. What do you observe about the data?

::: {.callout-tip collapse="true"}
## Expand to see solution

```{r}
ggplot(data=b, aes(x=ind, y=t, color=trait)) + 
  geom_point() + 
  facet_grid(~ geno)
```
:::

## Always plot your data

```{r}
library(tidyverse)
d <- read_tsv("data/example.tsv")
n1 <- rep(c("x","y"), 13)
n2 <- c("","",rep("_",24))
n3 <- c("", "", c(sort(rep(c(1:12), 2))))
names(d) <- paste0(n1,n2,n3)
names(d)
```

## Similar regression lines

These three data sets have very similar regression lines:

```{r}
summary(lm(x ~ y, data=d)) %>%  coef()
summary(lm(x_1 ~ y_1, data=d)) %>%  coef()
summary(lm(x_3 ~ y_3, data=d)) %>%  coef()
```

```{r,message=FALSE}
ggplot(d,aes(x=x,y=y)) + geom_point() +
  geom_smooth(method="lm") + ggtitle("Linear regression of y ~ x")
```

Now try this:

```         
ggplot(d,aes(x=x_1,y=y_1)) + geom_point() + 
  geom_smooth(method="lm")
```

::: {.callout-tip collapse="true"}
## Expand to see solution

```{r}
ggplot(d,aes(x=x_1,y=y_1)) + geom_point() + 
  geom_smooth(method="lm") + ggtitle("Linear regression of y_1 ~ x_1")
```
:::

And now try this:

```         
ggplot(d,aes(x=x_3,y=y_3)) + geom_point() + 
  geom_smooth(method="lm")
```

::: {.callout-tip collapse="true"}
## Expand to see solution

### Always plot your data!

```{r}
ggplot(d,aes(x=x_3,y=y_3)) + geom_point() + 
  geom_smooth(method="lm") + ggtitle("Linear regression of y_3 ~ x_3")
```
:::

## Always plot your data

```{r message=FALSE,warning=FALSE}
f <- read_tsv("data/BoxPlots.tsv")
# Delete the first column
f <- f[,-1]
head(f)
```

Stacking vectors concatenates multiple vectors into a single vector along with a factor indicating where each observation originated.

```{r}
head(stack(f),2)
```

Now try this:

```         
ggplot(stack(f), aes(x = ind, y = values)) +
  geom_boxplot()
```

::: {.callout-tip collapse="true"}
## Expand to see solution

## Identical box plots

These data have essentially identical box plots.

```{r echo=FALSE}
ggplot(stack(f), aes(x = ind, y = values)) +
  geom_boxplot()
```
:::

## Boxplots

While the box plots are identical, box plots may not tell the whole story.

Let's try violin plots instead:

```         
ggplot(stack(f), aes(x = ind, y = values)) +
  geom_violin()
```

A violin plot is a mirrored density plot.

::: {.callout-tip collapse="true"}
## Expand to see solution

## Non-identical violin plots

```{r echo=FALSE}
ggplot(stack(f), aes(x = ind, y = values)) +
  geom_violin()
```
:::

## Sina plots

Sidiropoulos N, Sohi SH, Pedersen TL, Porse BT, Winther O, Rapin N, Bagger FO. SinaPlot: An Enhanced Chart for Simple and Truthful Representation of Single Observations Over Multiple Classes. Journal of Computational and Graphical Statistics. Taylor & Francis; 2018 Jul 3;27(3):673–676. DOI: <https://doi.org/10.1080/10618600.2017.1366914>


```         
library(ggforce)
ggplot(stack(f), aes(x = ind, y = values)) +
   geom_violin() + geom_sina()
```

::: {.callout-tip collapse="true"}
## Expand to see solution

## Sina plots

```{r echo=FALSE}
library(ggforce)
ggplot(stack(f), aes(x = ind, y = values)) +
   geom_violin() + geom_sina()
```
:::

## Sina plots

`method == "counts"`: The borders are defined by the number of samples that occupy the same bin.

```         
ggplot(stack(f), aes(x = ind, y = values)) +
   geom_violin() + geom_sina(method="count")
```

::: {.callout-tip collapse="true"}
## Expand to see solution

## Sina plots

```{r echo=FALSE}
ggplot(stack(f), aes(x = ind, y = values)) +
   geom_violin() + geom_sina(method="count")
```
:::

## Drawing multiple graphs

Sometimes we'd like to draw multiple plots, looping across variables. Doing this within an R Markdown or Quarto Markdown document using `ggplot2` is tricky. See <https://dplyr.tidyverse.org/articles/programming.html> and <https://r4ds.hadley.nz/functions.html#plot-functions> for details.

Here's one way to do this - this example code will generate two scatter plots:

```{r}
x.names <- c("x","x_1")
y.names <- c("y", "y_1")
for (i in 1:2) {
  x.nam <- sym(x.names[i])
  y.nam <- sym(y.names[i])
  print(ggplot(data=d, aes(x = {{ x.nam }}, 
                           y = {{ y.nam }})) + 
          geom_point())
}
```

## Writing ggplot functions

See <https://r4ds.hadley.nz/functions.html#plot-functions>

```{r}
library(palmerpenguins)

PlDensity <- function(fill, ...) {
  ggplot(penguins %>% filter(!is.na(bill_length_mm)), 
         aes(bill_length_mm, fill = {{ fill }})) +
    geom_density(alpha = 0.5) +
    facet_wrap(vars(...))
}
```

Example from: <https://twitter.com/yutannihilat_en/status/1574387230025875457?s=20&t=FLbwErwEKQKWtKIGufDLIQ>

```{r}
PlDensity(species)
```

```{r}
PlDensity(island, sex) %>% print() %>% suppressWarnings()
```

```{r}
PlDensity(sex, island, year) %>% print() %>% suppressWarnings()
```

## Exercise 3

Consider this example code:

```{r}
histogram <- function(df, var, binwidth) {
  df |> 
    ggplot(aes({{ var }})) + 
    geom_histogram(binwidth = binwidth)
}
```

From: <https://twitter.com/hadleywickham/status/1574373127349575680?s=20&t=FLbwErwEKQKWtKIGufDLIQ>

When applied to the quantitative trait `t` from the data frame `b`, this generates this histogram:

```{r}
histogram(b, t, 0.1)
```

### Exercise

After reading the example above, extend the `histogram` function to allow facetting and use it to draw a histogram of the quantitative trait `t` facetted by `geno` using the data set `b` that we set up above.

::: callout-tip
## Hints

-   See <https://r4ds.hadley.nz/functions.html#plot-functions>
-   Use the `{{ var }}` approach
:::

::: {.callout-tip collapse="true"}
## Expand to see solution

Hadley Wickham states:

You have to use the vars() syntax

```         
foo <- function(x) {
  ggplot(mtcars) +
    aes(x = mpg, y = disp) +
    geom_point() +
    facet_wrap(vars({{ x }}))
}
```

Tweet: <https://twitter.com/hadleywickham/status/1574380137524887554?s=20&t=FLbwErwEKQKWtKIGufDLIQ>

```{r}
histogram <- function(df, var, binwidth, grp) {
  df |> 
    ggplot(aes({{ var }})) + 
    geom_histogram(binwidth = binwidth) +
    facet_wrap(vars({{ grp }}))
}
```

```{r}
histogram(b, t, 0.1)
```

```{r}
histogram(b, t, 0.1, geno)
```
:::

## Source of data

Illustrative data sets from <https://www.research.autodesk.com/publications/same-stats-different-graphs/>
