# Lecture: R Basics

## R Basics lecture

Here's a recording of this lecture (48 minutes 14 seconds):

[Recording](https://pitt.hosted.panopto.com/Panopto/Pages/Viewer.aspx?id=33102882-3b59-4c1f-898b-b07400ea6d7b){.external target="_blank"}

<iframe credentialless src="https://pitt.hosted.panopto.com/Panopto/Pages/Embed.aspx?id=33102882-3b59-4c1f-898b-b07400ea6d7b&amp;autoplay=false&amp;offerviewer=true&amp;showtitle=true&amp;showbrand=true&amp;captions=false&amp;interactivity=all" height="405" width="720" style="border: 1px solid #464646;" allowfullscreen allow="autoplay">

</iframe>

## R Basics slides

[PDF slide set](pdfs/R_basics.pdf){target="_blank"}

<iframe width="100%" height="800" src="pdfs/R_basics.pdf">
</iframe>

## Questions

### Question 1

What does `a <- 9; a < - 9` do?

```{r, echo=FALSE}
library(webexercises)
choices = c(
"TRUE",
"9",
answer = "FALSE",
"-9"
)
```

`r longmcq(choices)`


### Question 2

Given a vector `x` of length 10, which of the following R code snippets would remove the third element from the vector and assign the result to a new vector `y`?  More than one option may be correct.

    
```{r, echo=FALSE}
choices = c(
answer = "y <- x[-3]",
"y <- x[3] <- NULL",
answer = "y <- x[c(1:2, 4:10)]",
"y <- x[1:2, 4:10]"
)
```

`r longmcq(choices)`

### Question 3

Given the following vector:

`x <- c(a = 1, b = 2, c = 3)`

Which of the following R code snippets will extract only the element corresponding to the name "b"?

```{r, echo=FALSE}
choices = c(
answer = "x[2]",
answer = 'x["b"]',
'x[x == "b"]',
"x[x > 1]"
)
```

`r longmcq(choices)`    

### Question 4

Given the following R code:

```
x <- 1:6
dim(x) <- c(2, 3)
```

Which of the following options correctly extracts the second column of the matrix `x` as a column vector?

```{r, echo=FALSE}
choices = c(
"x[2, ]",
'x[, 2]',
'x[2]',
answer = "x[, 2, drop = FALSE]"
)
```

`r longmcq(choices)`  


### Question 5

Which of the following statements are true about data frames in R?

```{r, echo=FALSE}
choices = c(
"A data frame is a type of list where all elements must be of the same data type.",
answer = 'A data frame is a list where elements can be of different data types.',
'A data frame is a matrix where elements must be of the same data type.',
answer = "A data frame is a list where all elements of the list must be of the same length."
)
```

`r longmcq(choices)`      