---
editor: 
  mode: source  
---  

# Lecture: R: factors, subscripting

## R: factors, subscripting lecture

Here's a recording of this lecture (43 minutes 25 seconds):

[Recording](https://pitt.hosted.panopto.com/Panopto/Pages/Viewer.aspx?id=850e5bbc-cbad-4d5d-967e-b06f00fa2c54){.external target="_blank"}

<iframe credentialless src="https://pitt.hosted.panopto.com/Panopto/Pages/Embed.aspx?id=850e5bbc-cbad-4d5d-967e-b06f00fa2c54&amp;autoplay=false&amp;offerviewer=true&amp;showtitle=true&amp;showbrand=true&amp;captions=true&amp;interactivity=all" height="405" width="720" style="border: 1px solid #464646;" allowfullscreen allow="autoplay">

</iframe>

## R: factors, subscripting slides

[PDF slide set](pdfs/R_factors_dates.pdf){target="_blank"}

<iframe width="100%" height="800" src="pdfs/R_factors_dates.pdf">
</iframe>

## Questions

### Question 1

Which of the following statements about factors in R is correct?

```{r, echo=FALSE}
library(webexercises)
choices = c(
"Factors can only use values that are among the initially specified levels, and cannot have new levels added later.",
"Factors are primarily used for numerical calculations rather than categorical data.",
answer = "Ordinal data can be coded using ordinal factors to represent the order of categories.",
"Factors can be used to add more understandable labels to data, but they do not support ordinal data."
)
```

`r longmcq(choices)`


### Question 2

Which of the following R code snippets correctly demonstrates valid methods for subsetting data?

```{r, echo=FALSE}
choices = c(
answer='x[2], x["name"], x[x > 5]',
answer='x[2], x["name"], x[grep("pattern", x)]',
answer = 'x[2], x["name"], x[!is.na(x)]',
'x["name"], x[TRUE], x[order(x)]'
)
```

`r longmcq(choices)`


### Question 3

Which of the following R code snippets best illustrates how NA values can interfere with subsetting, where x is c(1, 6, NA, 8)?

```{r, echo=FALSE}
choices = c(
'x[!is.na(x)]',
'x[2]',
'x[1:3]',
answer = 'x[x > 5]'
)
```

`r longmcq(choices)`
