# Unix Miscellaneous

## Unix Miscellaneous

Here's a recording of this lecture (19 minutes 57 seconds):

[Recording](https://pitt.hosted.panopto.com/Panopto/Pages/Viewer.aspx?id=93c2f294-2351-4a20-92f1-b1fe010fb277){.external target="_blank"}

<iframe credentialless src="https://pitt.hosted.panopto.com/Panopto/Pages/Embed.aspx?id=93c2f294-2351-4a20-92f1-b1fe010fb277&amp;autoplay=false&amp;offerviewer=true&amp;showtitle=true&amp;showbrand=true&amp;captions=true&amp;interactivity=all" height="405" width="720" style="border: 1px solid #464646;" allowfullscreen allow="autoplay">

</iframe>

## Unix Basics slides

[PDF slide set](pdfs/unix_miscellaneous.pdf){target="_blank"}

<iframe width="100%" height="800" src="pdfs/unix_miscellaneous.pdf">

</iframe>

## Questions

### Question 1

What command would make a symbolic link to /really/long/path/to/data/folder/ and name it data_shortcut? What command would remove the link?

```{r, echo=FALSE}
library(webexercises)
choices = c(
"ln -s /really/long/path/to/data/folder/ data_shortcut ; rm data_shortcut",
answer="ln -s /really/long/path/to/data/folder/ data_shortcut ; unlink data_shortcut",
"ln -s data_shortcut /really/long/path/to/data/folder/  ; rm data_shortcut",
"ln -s data_shortcut /really/long/path/to/data/folder/ data_shortcut ; unlink data_shortcut"
)
```

`r longmcq(choices)`

### Question 2

Which of the following are needed/advised when running an R script on htc? Multiple answers may be correct.

```{r, echo=FALSE}
library(webexercises)
choices = c(
answer="Loading the modules you need (a version of gcc and a version of R)",
answer="Running the script with a command like Rscript --vanilla my_script.R",
answer="Including options(echo=TRUE) at the top of the R script so that each line of R code is echoed before its output",
answer="Doing args <- (commandArgs(T)) and then for (i in 1:length(args)) { eval(parse(text =args[[i]])) } in the script to parse command-line arguments"
)
```

`r longmcq(choices)`

### Question 3

Which of the following are true of Slurm job arrays?

```{r, echo=FALSE}
library(webexercises)
choices = c(
answer="They are an alternative to using a loop to repeat similar tasks.",
answer="They implement parallelism by running each task in the array as a separate Slurm job.",
answer="They are slightly simpler to use when the tasks/files you need to loop over are labelled 1, 2, 3, ... n.",
answer="Add #SBATCH --array=1-n to your Slurm script in order implement the array (replacing n with the number of tasks in the array).",
answer="Inside the script, the variable $SLURM_ARRAY_TASK_ID functions like a looping index variable that you can use to control which task gets executed in each job in the array.",
answer="However, they can still loop over any list of files by using syntax like that shown in the last slide."
)
```

`r longmcq(choices)`