# Unix Data Manipulation

## Unix Data Manipulation

Here's a recording of this lecture (69 minutes 22 seconds):

[Recording](https://pitt.hosted.panopto.com/Panopto/Pages/Viewer.aspx?id=12821816-9291-498a-b965-b1fd01281ecc){.external target="_blank"}

<iframe credentialless src="https://pitt.hosted.panopto.com/Panopto/Pages/Embed.aspx?id=12821816-9291-498a-b965-b1fd01281ecc&amp;autoplay=false&amp;offerviewer=true&amp;showtitle=true&amp;showbrand=true&amp;captions=true&amp;interactivity=all" height="405" width="720" style="border: 1px solid #464646;" allowfullscreen allow="autoplay">

</iframe>

## Unix Basics slides

[PDF slide set](pdfs/unix_data_manipulation.pdf){target="_blank"}

<iframe width="100%" height="800" src="pdfs/unix_data_manipulation.pdf">

</iframe>

## Questions

### Question 1

Suppose file.bim is a tab-delimited plink bim file (meaning the second column has rsIDs). Which command will find the number of lines corresponding to rs2345, without finding spurious matches like rs23456?

Please select the correct options (more than one may be correct).

```{r, echo=FALSE}
library(webexercises)
choices = c(
"grep rs2345 file.bim",
"grep -o rs2345 file.bim",
"grep -oc rs2345 file.bim",
answer="grep -wc rs2345 file.bim",
answer="grep -w rs2345 file.bim | wc -l",
answer="awk '$2 == \"rs2345\"' file.bim | wc -l",
answer='grep -c "\brs2345\b" file.bim (the unprinted characters are supposed to be word boundaries, i.e., backslash-b)',
answer='cut -f2 file.bim | sort | uniq -c | awk \'$2 == "rs2345" {print $1}\''

)
```

`r longmcq(choices)`

### Question 2

Which of the following cut commands extracts the first and third columns (in that order) from a comma-separated file named data.csv?

Please select the correct option(s).

```{r, echo=FALSE}
library(webexercises)
choices = c(
answer="cut -d"," -f1,3 data.csv",
answer="cut -d"," -f3,1 data.csv",
"cut -d"," -f1-3 data.csv",
answer='awk -F"," \'{print $1"\t"$3}\' data.csv'
)
```

`r longmcq(choices)`

### Question 3

Which of the following commands correctly outputs a count of the unique lines from a file named records.txt, sorted by the most frequent line first?

Please select the correct option(s).

```{r, echo=FALSE}
library(webexercises)
choices = c(
"uniq -c records.txt | sort -k1,1nr",
"sort -u records.txt | uniq -c | sort -k1,1nr",
answer="sort records.txt | uniq -c | sort -k1,1nr",
"sort -c records.txt | uniq -d | sort -nr"
)
```

`r longmcq(choices)`

### Question 4

Which of the following statements about awk is correct?

Please select the correct option(s).

```{r, echo=FALSE}
library(webexercises)
choices = c(
"awk 'print $1' file prints the first column of file.",
answer="awk '{ print $1 }' file prints the first column of file.",
answer="awk '{ print $2, $3 }' file prints the second and third columns, separated by a space.",
answer='awk \'{ print $2 "\t" $3 }\' file prints the second and third columns, separated by a tab.'
)
```

`r longmcq(choices)`

### Question 5

Which sed command will recode chromosome X as chromosome 23 in a plink bim file?

Please select the correct option(s).

```{r, echo=FALSE}
library(webexercises)
choices = c(
"sed 's/^X$/23/' file.bim",
answer="sed 's/^X/23/' file.bim",
"sed 's/X/23/' file.bim",
"sed '/^X/d' file.bim"
)
```

`r longmcq(choices)`

### Question 6

Which command would flip a DNA strand to its complement?

Please select the correct option(s).

```{r, echo=FALSE}
library(webexercises)
choices = c(
"tr 'AT' | tr 'CG' | tr 'GC' | 'TA'",
"tr 'ACGT' 'TG'",
"tr 'AGTC' 'TCGA'",
answer="tr 'ACGT' 'TGCA'"
)
```

`r longmcq(choices)`